\documentclass[12pt]{article}

\usepackage[a4paper,lmargin=25pt,rmargin=25pt,bmargin=0.8cm,includefoot]{geometry}
\usepackage[T1]{fontenc}
\usepackage{parskip}
\usepackage{microtype}
\usepackage{marginnote}
\usepackage{fwlw}
\usepackage{librebaskerville}

\pagestyle{NextWordFoot}
\reversemarginpar
\renewcommand*\marginfont{\footnotesize}

\begin{document}

\section{Introduction}

\large{
To all but the dinosaurs, meteors are quite cool. For my EMC project I have
studied 6 different areas of radio meteor detection, with the aim of analysing
variation and phenomena associated with this method of detection. Before I
present these studies, we must consider what a meteor {\it is}.  
}

\section{What is a meteor?}

\large{
I must admit that I have lied to all of you: only half of meteors rock. The
other half are metallic in composition. It was a catchy title though, I think.\\

Meteors have three stages to their lives. First, they exist as small rocky or
metallic bodies floating through space, where they are referred to as
meteoroids. When, or if, a meteoroid enters the atmosphere of Earth, it becomes
a meteor. As re-entry occurs, the meteor is heated intensely, owing to friction
with air molecules. The air around the meteor is also heated, to such an extreme
temperature that atoms making up the air are ionised, stripping electrons away
from them. The super-heated trail left by the meteor is what you see as a
`shooting star'. If the meteor makes it to the ground, it becomes a meteorite.
}

\section{Meteor showers}

\large{
Typically, meteors are observed as part of `showers'. These are events, lasting
over a couple of weeks, where there is an increased number of meteoroids in the
space around Earth, resulting in more meteors, all of which appear to originate
from a single point in the sky, called the radiant. Meteors that are not part of
a shower are referred to as `sporadic' meteors.\\

Meteor showers occur throughout the year, and are each attributed to a comet, or
sometimes an asteroid, which causes the shower. As the comet orbits the Sun, a
trail of dust is left, producing a meteor `stream'. The Earth moving through
this stream is what constitutes a meteor shower. When Earth moves through the
most dense part of the stream, many more meteors are observed. This is known as
the `peak' of the shower.\\

The names for the showers are derived from the constellation where the radiant
lies. For example, the Perseids shower's radiant is in Perseus. Each shower
occurs at roughly the same time each year, since the trail left by the source
comet doesn't move relative to Earth, meaning it lies at the same point in
Earth's orbit each year, occurring at the same time. \\

These showers can vary widely. Some may produce lots of fireballs, which are
meteors that explode under the pressure of re-entry, lighting up the entire sky.
Some may only produce very small meteors, since there is only small sized debris
left by a comet. This variation occurs from the meteor `stream' spreading out as
the trail orbits the Sun. As well as this, over successive years of meteor
showers without another orbit of a comet, the larger debris is collected by
Earth and only smaller debris is left. This can result in peak years, when many
more meteors are detected, after an orbit of the source comet. Conversely, this
can also cause the peak of a shower to vary in intensity over a number of years.
}

\section{Radio Meteor Detection}

\large{
Radio meteor detection is possible because the ionised, super-heated air left as
a trail by a meteor is reflective to radio waves. The process starts with a
transmitting antenna, often a significant distance away from those observing the
meteors, which emits a radio signal into the atmosphere. This signal must
ideally spread across a wide area of sky, allowing the best chance of detecting
a meteor. The signal must also be at a suitable angle such that the signal
intercepts the meteors where most re-entries occur: between 60 and 110 km.\\

When the signal intercepts the meteor trail, it is reflected and the signal can
then be received by an observer. The part of a received signal that indicates a
meteor detection is referred to as an `echo'. Often the transmitter will be
beyond the horizon of the receiver, allowing no direct signal, which is best for
meteor detection, since the received echo is clearer.\\

What can we do with a meteor echo? Well, there are two main methods: simply
counting the number of detections an hour, or capturing images of echoes for
further analysis. Each of these methods depends on a precursor, where the
received signal is represented as a `waterfall' plot. These are plots that show
frequency and intensity over time. The frequency shown in the plot is not
necessarily accurate due to the way the radio signal is received. The intensity,
in dB, is represented using a colour scale. Meteor echoes have a distinctive
look in a waterfall plot. \\

Using this waterfall plot, we can complete the two methods. For counting the
number of detections an hour, we analyse the signal, and declare any peaks above
a certain intensity threshold a meteor. In order to capture images, we generally
use a 2D representation of the waterfall plot. \\
}

\section{Data}
\large{
In my analyses, I have used data from the Radio Meteor Observation Bulletin
organisation, known as RMOB, which provides a database of hourly counts from
observers around the world. The majority of observers contributing data are 
amateurs, and have various set-ups.

\section{Diurnal Shift}

\large{
Diurnal shift, also known as diurnal variation, is a phenomenon exhibiting an
increase and decrease in meteor detection counts over the course of a day. This
is a well documented observation, appearing to occur across the globe. Through
my own research I found no standardised model of what causes diurnal shift,
though there appears to be an accepted explanation for the event.\\

Based on existing explanations, I made my own model, an extension and more
mathematical description of that already given, which is based on the idea that
the overall intercept velocities of sporadic meteors changes due to the Earth's
orbital velocity. This is better illustrated with a diagram. \\

We must start with the assumptions that sporadic meteors have no preferred
incident direction. We also assume that the number of meteors detected is
proportional to the intercept velocity of any given meteor. Meteors are mostly
only detected if they are travelling tangent to Earth's atmosphere, which
complicates things. However, all of these velocities will cancel, leaving an
overall incident velocity perpendicular to Earth's surface. From the diagram we
see that the incident velocity is the addition of a component of Earth's orbital
velocity, and the meteor's velocity. At sunset, the meteor's incident velocity
is added to Earth's orbital velocity, giving a higher intercept velocity and
thus more meteors are detected.\\ 

In order to test my model, I used the RMOB data to analyse the diurnal variation
of each observer. Considering the form of diurnal shift for different locations,
we see that the model fits well with the data: plots of diurnal shift for
observers in Europe, North America, as well as Asia \& Australia all produce
diurnal variation fitting a sinusoidal curve, as predicted by the model.\\

The model also predicts that the peak hour will be approximately 6am. Only
approximately since nature is unfortunately not as well behaved as we'd like.
This map shows each observer as a dot, with the colour based on the peak hour of
their diurnal shift. It is somewhat clear that the peak hour is based on their
longitude. Correcting for each observer's timezone (based on longitude, not {\it
actual} timezones), we see that the majority of observers have a peak hour of
6am, which agrees with the model.\\

Although rather hard to discern, my results also indicate that the peak hour of
diurnal shift does not change over time, nor the intensity. However, the
sinusoidal fit of an observer's diurnal variation appears to be less pronounced
between 2005 and 2011. We'll revisit the reason for this later...\\
}

\section{Spatial Variation}

\large{
Analysing the spatial variation involves studying the change in a number of data
characteristics in relation to longitude and latitude. The aim of this was to
investigate whether there are more suitable locations for radio meteor
detections. The short answer is: no. Generally, longitude and latitude do not
appear to influence the quality or quantity of an observer's data.\\

The long answer is that, in order to describe, or summarise, an observer's
dataset, we use 6 `data characteristics'. The first are your typical mean and
maximum. Minimum counts are a somewhat useless measure, since the minimum count
will almost always be 0. As well as the maximum and mean, the standard error for
all counts for a single month is taken. The standard error is a measure of how
much the data varies, with consideration of how large the data set is. This
measure indicates whether the observer records consistent counts or otherwise.
Skew is an indicator of the shape of a set of data. Recording the skew of a
given month of data from an observer indicates whether there is a clear limit to
the detection counts, or the counts trail off towards larger numbers. Finally,
we calculate the fit of the given data to a sine curve, indicating how clear
diurnal shift is in the data, including the peak hour of this diurnal shift.\\

Using these results, we can see how they change based on latitude and longitude.
As I have said, the results are inconclusive. The only clear correlation is
between longitude and peak hour of diurnal shift, though this is expected from
the model we have discussed. A significant problem in this analysis was sample
sizes: with only 4 or 5 observers in each location category, conclusive results
are hard to come by.
}

\section{Temporal Variation}

\large{
In addition to analysing the change of the data characteristics through space, I
analysed their change through time, on three scales: variation since 2000,
variation over a year, and variation over a month.\\

Regarding the month scale, there is no clear change in detection count for any
observer. There is a very slight increase, but only on the order of 4 or 5
counts - this is not significant, though the change is clear in each location
category. Very small increases such as this are most likely caused by a large
meteor shower occuring on the same day of the month each year, such as the
Perseids, which peak between the 12th and 14th of August each year.\\

I found that, on the scale of a single year, the detection counts are lower from
January until May. Beyond and including June, the detection counts are larger by
around 20. The peak of this increase is in June. Peculiarly this variation
doesn't appear to apply to the Asia \& Australia category. However, there is
still an increase of a similar magnitude, simply peaking two months earlier.
This increase could be caused by several factors. It could be due to the timings
of meteor showers, or potentially the position of planets relative to Earth. For
example, Jupiter often `sweeps' up debris in space, preventing it reaching
Earth. However, this is unlikely, since the variation is periodic and, although
the positioning of planets is also periodic, it is with a very different
frequency. In order to investigate this further, factoring out the background
number of detection counts would be necessary, as well as analysing the change
in period of variation over time.\\

To my surprise, I found a significant change in detection counts since 2000.
Between 2005 and 2011, the detection counts increase up to almost 2 times the
normal value. Upon further research, I found this was a documented phenomenon.
Other studies have noted the correlation between the maximum of the detection
counts and minimum of the solar cycle. This is a reasonable link: when the Sun
is less active, less solar wind is generated, giving a lower electron density in
the upper atmosphere of Earth, which itself provides less interference to radio
signals, allowing a greater number of detections.\\

In addition to this, I have found that there is no significant difference
between detection counts during the day compared to the night. This demonstrates
that facing towards or away from the Sun has no influence on how many meteors
are detected. In addition to this, the results suggest that the distribution of
sporadic meteors is relatively uniform in space, giving validity to assumptions
made for the diurnal shift model.\\
}

\section{Antenna Type}

\large{
My final analysis of variation focuses on the difference in data collected by
observers with different types of antenna. Overall there are 9 clear antenna
types, with some miscellaneous types. Overwhelmingly the most popular is the
Yagi antenna. I used the same data characteristics in order to discern
differences between the types.\\

In regards to the mean and maximum counts, most antenna types had the same
results. The most important result of this characteristic, and those that
follow, is the large error in each value. The implication of this large error is
how much difference there is between data from observers with the same
antenna. In fact, for all results, the values didn't vary much, and the errors
are large. It's clear that the antenna being used doesn't influence the data
collected in any influential way.\\
}

\section{Zenithal Hourly Rate}

\large{
It is important to have a standard way of measuring the activity of a meteor
shower, enabling predictions to be made by models of debris streams. This
measure is called the Zenithal Hourly Rate (ZHR). ZHR indicates how many meteors
an observer can expect to see in an hour if they can see the entire sky, with
perfect viewing conditions, and the radiant is directly overhead, called the
zenith. The formula for calculating the ZHR uses certain correction factors to
compensate for {\it not} having these conditions. For example, there is a
correction for the height of the radiant above the horizon, the field of view,
and the limiting magnitude, which is the dimmest magnitude an object you can
see.\\

To start my analysis, I modified the formula so that it works. The correction
factor for radiant height is inaccurate, indicating that no meteors are seen
when the radiant is below the horizon, which is not the case. It also results in
an infinite ZHR when the radiant approaches the horizon. I formulated my own
correction factor, and have used that instead. In order to use the formula for
radio meteor detection, other factors must be changed. For starters, the field
of view correction factor cannot be used since it is not known exactly how much
of the sky each antenna can see. Unfortunately, beam patterns are not helpful
either, since it is dependent on how much of the sky can be seen by both the
receiving and transmitting station.\\

As well as this, the correction factor for limiting magnitude is invalid. This
factor uses the idea of population index, which indicates how many times more
meteors you expect to see with an increase in magnitude of 1. However, this
assumes that there is an exponential increase up to the visual limiting
magnitude of 6.5, but this cannot possibly be valid for the limiting magnitude
of a radio antenna which is {\it much} greater. If it {\it was} the case, we
would expect to see trillions of meteors an hour -- and we definitely don't.\\

Using the formula for radio observation also requires a correction for the
background detection rate. This is not applicable to visual observation since
the background detection rate is close enough to 0 to be negligible. However,
even when there are no active showers, many meteors are detected by radio
observation. When these correction factors are included, the formula is like so.
This comes with the unfortunate assumptions that the transmitting antenna can
see the entire sky.\\

Despite these assumptions, the results correlate well with records of visual
maximum ZHRs and expected shower ZHRs. There is a clear offset between the
results, which would most likely be accounted for by knowing the limiting
magnitude and field of view of each respective radio observation. The results
demonstrate that ZHR is an applicable measure of activity through radio
observation, though there are of course improvements to be made.\\
}

\section{Root Mean Square Difference Image Analysis}

\large{
As mentioned at the start of this talk, the main method of radio meteor
detection is to count the number of meteors detected in an hour. Also mentioned
is the alternative of recording images of waterfall plots. In an attempt to use
this as a quantitative measure of meteor flux (a fancy word for flow, i.e. the
number of meteors coming through Earth's atmosphere), I wrote a program that
uses Root Mean Square Difference.\\

The idea of using RMS Difference lies in how meteor echoes are represented in
waterfall plots. Since the most intense signals, indicating meteors, are white
in colour, which has the largest colour value when represented in an image. RMS
difference is the square root of the mean of the squared differences between
corresponding pixels in two images. Effectively, it's a measure of how different
two images are. By comparing a waterfall plot to a baseline image -- the choice
of baseline is arbitrary -- we gain a value for how different two images are.
Considering the change in this value over time indicates the change in meteor
flux.

Using separately recorded hourly counts, from the same observation station
(which is the Norman Lockyer Observatory), we see that the correlation between
my own results and the detection counts are good. The correlation coefficient of
these two sets of data is 0.616, indicating a good positive correlation. That
is, if one increases, the other does so accordingly. This provides a new way of
analysing meteor detection results, which has no cost on top of detection
equipment and only requires one program to use. 
}

\section{Conclusion}

\large{
To conclude, more data is needed! A more uniform coverage of the globe, and more
data, would dispel much of the doubt over results with low sample sizes.
However, saying this, I have found some interesting results, which pose
significant questions and more investigation is always needed.
}

\end{document}
